---
title: "Notebook 3: bandle"
author: "Olivia J. Conway"
date: "`r Sys.Date()`"
fontsize: 11pt
output:
  rmdformats::robobook:
    theme: default
    fig_caption: yes
    use_bookdown: yes
pdf_document:
    toc: yes
    number_sections: true
editor_options:
  chunk_output_type: console
---

# Overview
This notebook contains the R code used to perform differential localisation 
analyses using the `bandle` package.

## Install and load packages

Install R packages needed for data processing -

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
install.packages("BiocManager")
}

BiocManager::install(c("bandle",
                       "coda",
                       "tictoc",
                       "here"))
```

Load packages

```{r, eval=TRUE,message=FALSE, warning=FALSE}
library("bandle")
library("coda")
library("tictoc")
library("here")
```

## Load data

```{r, eval=TRUE}
# individual replicates (basal_prots, insulin_prots)
load(here("output/adipocyte_bandle_msnsets.rda"))
```

## The `bandle` package

In this analyses we use the [`bandle` package](https://www.bioconductor.org/packages/release/bioc/html/bandle.html) to perform differential
localisation analyses. `bandle` which stands for Bayesian ANalysis of
Differential Localisation, is a method developed by [Crook et al. in 2022](https://doi.org/10.1038/s41467-022-33570-9) which
implements integrative semi-supervised functional mixture model, to obtain the
probability of a protein being differentially localised between two conditions.

# Preparing data for bandle

## Data formatting

`bandle` requires two `lists` of `MSnSets` one for each condition, containing
the common proteins between all conditions and replicates

```{r, eval=TRUE}
## put the MSnSets into a "list"
quant_list <- list(basal_prots[[1]], basal_prots[[2]], basal_prots[[3]],
                   insulin_prots[[1]], insulin_prots[[2]], insulin_prots[[3]])

## Keep common proteins across all replicates and conditions
quant_list <- commonFeatureNames(quant_list)

## Put data into 2 lists one for each condition
basal <- list(quant_list[[1]], quant_list[[2]], quant_list[[3]])
insulin <- list(quant_list[[4]], quant_list[[5]], quant_list[[6]])
all_conditions <- c(basal, insulin)
```

## Fitting Gaussian Process
First we fit non-parametric regression functions to the markers profiles.  

<br>  

### Prior
The prior needs to form a `K*3` matrix (where `K` is the number of organelle classes in the data).  
One column is for the prior, one for the length-scale amplitude and one for the SD parameters. 
```{r, eval=TRUE}
# Find number of organelle classes in data
K <- length(getMarkerClasses(all_conditions[[1]], fcol = "markers"))
```

```{r, eval=TRUE}
pc_prior <- matrix(NA, ncol = 3, K)
pc_prior[seq.int(1:K), ] <- matrix(rep(c(1, 50, 50),
                                       each = K), ncol = 3)
```

### Fit GP using priors
Now the complexity priors have been generated, we pass them as an argument to the `fitGPmaternPC` function.  
```{r, fig.width=9, fig.height=16, eval=TRUE}
gpParams <- lapply(all_conditions, function(x) fitGPmaternPC(x,  hyppar = pc_prior))
```

Plot the GP fits
```{r, eval=TRUE, fig.width=9, fig.height=9}
## create labels for the plots
id <- c(paste0("basal-rep", 1:3), paste0("insulin-rep", 1:3))

## now plot all GPs
for (i in seq(all_conditions)) {
  # png(file = paste0("3_running_bandle/figures/gpParams_", 
  #                   id[i], ".png"), height = 4000, width = 4000, 
  #     units = "px", res = 400)
  par(mar=c(2,2,2,2), oma = c(2,2,6,2), mfrow = c(4,3)) 
  plotGPmatern(all_conditions[[i]], params = gpParams[[i]])
  mtext(id[i], side = 3, outer = TRUE, font = 2, cex = 2, line =2)
  # dev.off()
}
```

## Setting the prior on weights
Dirichlet distribution is a family of continuous multivariate probability
distributions parameterized by a vector of positive reals. It is a multivariate
generalization of the beta distribution.
<br>
```{r, eval=TRUE}
set.seed(1)
dirPrior = diag(rep(1, K)) + matrix(0.0002, nrow = K, ncol = K)

### sanity check, doesn't actually affect bandle function as only dirPrior goes in
predDirPrior <- prior_pred_dir(object = all_conditions[[1]],
                               dirPrior = dirPrior,
                               q = 15)
```


Save the parameters,

```{r, eval=FALSE}
save(basal, insulin, gpParams, pc_prior, dirPrior, file = here("output/adipocyte_lopit_bandle_input.rda"))
```

# Running BANDLE

The `bandle` method is computationally intensive requires a HPC to run. We used
the University of Cambridge HPC to run the analysis. The output `bandle` run 
has been uploaded to Zenodo at xxx with DOI (add URL and DOI).

```{r, eval=FALSE}
bandle_output <- bandle(objectCond1 = basal,
                        objectCond2 = insulin,
                        numIter = 30000,
                        burnin = 5000,
                        thin = 20,
                        gpParams = gpParams,
                        pcPrior = pc_prior,
                        numChains = 4,                       
                        dirPrior = dirPrior,
                        seed = 1)
```

