---
title: "Notebook 2: data visualisation"
author: "Olivia J. Conway"
date: "`r Sys.Date()`"
fontsize: 11pt
output:
  rmdformats::robobook:
    theme: default
    fig_caption: yes
    use_bookdown: yes
pdf_document:
    toc: yes
    number_sections: true
editor_options:
  chunk_output_type: console
---

# Overview
This script contains the R code used to visualise the LOPIT-DC data in "[paper title]".

## Install and load packages

Install R packages needed for data processing -

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
install.packages("BiocManager")
}

BiocManager::install(c("pRoloc",
                       "MSnbase",
                       "pRolocGUI",
                       "tidyverse",
                       "patchwork",
                       "gridExtra",
                       "colorspace",
                       "gplots",
                       "kableExtra",
                       "here",
                       "VennDiagram",
                       "eulerr",
                       "Rtsne",
                       "here"))
```

Load packages

```{r, eval=TRUE,message=FALSE, warning=FALSE}
library("pRoloc")
library("MSnbase")
library("pRolocGUI")
library("tidyverse")
library("patchwork")
library("gridExtra")
library("colorspace")
library("gplots")
library("kableExtra")
library("Rtsne")
library("here")
library("VennDiagram")
library("eulerr")
library("here")
```


Load functions
```{r, eval=TRUE}
source(here("R/prettymap.R"))
```

Load data
```{r, eval=TRUE, results = "hide"}
load(here("output/adipocyte_lopit_MSnSet.rda"), verbose = TRUE)
```

# Individual Replicates
## Summary

<!-- ```{r, eval=TRUE} -->
<!-- #Proteins in each rep in basal -->
<!-- lapply(basal_prots, dim) -->

<!-- #Proteins in each rep in insulin -->
<!-- lapply(insulin_prots, dim) -->
<!-- ``` -->

```{r, eval=TRUE}
tot_prots <- data.frame("Basal" = c(`Replicate 1` = nrow(basal_prots[[1]]), 
                                           `Replicate 2` = nrow(basal_prots[[2]]), 
                                           `Replicate 3`= nrow(basal_prots[[3]])),
                        "Insulin" = c(`Replicate 1` = nrow(insulin_prots[[1]]),
                                  `Replicate 2` = nrow(insulin_prots[[2]]),
                                  `Replicate 3` = nrow(insulin_prots[[3]])))
```

```{r, eval=TRUE, include=FALSE}
save(tot_prots,
     file = here("2_data_visualisation/output/tot_prots.rda"))
```

```{r kableoutput, warning=FALSE, message=FALSE, echo=FALSE}
kable(tot_prots, caption = "Number of quantified proteins per replicate per condition")
```

## Add markers

In this code chunk we read in the curated marker list from a .csv file

```{r, eval=TRUE}
mrk_df <- read.csv(here("data/markers.csv"), row.names = 1)
mrk <- mrk_df[,1]
names(mrk) <- rownames(mrk_df)
```

Now we add these markers to the data by matching the protein accession numbers
of the markers to the protein accession numbers in the data.

```{r, eval=TRUE}
basal_prots <- sapply(basal_prots, addMarkers, markers = mrk)
insulin_prots <- sapply(insulin_prots, addMarkers, markers = mrk)
```

Save this data for bandle run.

```{r, eval=TRUE}
save(basal_prots, insulin_prots, 
     file = here("output/adipocyte_bandle_msnsets.rda"))
```

## PCA plots

Generate principal components plots of the data

```{r, eval=TRUE}
pca_basal_prots <- lapply(basal_prots, plot2D, fcol = "markers", 
                    method = "PCA", plot = FALSE)
pca_insulin_prots <- lapply(insulin_prots, plot2D, fcol = "markers",
                    method = "PCA", plot = FALSE)

# mirror Y axis for 3rd rep of insulin only so all maps are the same orientation
pca_insulin_prots[[3]] <- plot2D(insulin_prots[[3]], fcol = "markers", 
                                 method = "PCA", plot = FALSE, mirrorY = TRUE)
```


```{r, eval=TRUE, fig.width=9, fig.height=9}
# png(file = here("2_data_visualisation/figures/Markers_PCA_replicates.png"), height = 5000, width = 5000, units = "px", res = 500)

## Set colours and parameters for plots
orgclasses <- getMarkerClasses(basal_prots[[1]])
mycol <- getStockcol()[seq(orgclasses)]
myleg <- c(orgclasses, "Unknown")
mycolleg <- c(mycol, "grey")
  
par(mfrow = c(3, 3))
par(mar = c(5, 5, 4, 3))

## Basal map
for (i in seq_along(basal_prots)) {
  prettyMap(
    pca_basal_prots[[i]], 
    basal_prots[[i]], 
    fcol = "markers", 
    ylim = c(-5, 4), xlim = c(-5, 6),
    main = paste0("Basal - replicate ", i, " \n(", 
                  nrow(basal_prots[[i]]), " proteins)"),
    mainCol = paste0(mycol, 70),
    outlineCol = darken(mycol))
  }

## Insulin map
for (i in seq_along(insulin_prots)) {
  prettyMap(
    pca_insulin_prots[[i]], 
    insulin_prots[[i]], 
    fcol = "markers", 
    ylim = c(-5, 4), xlim = c(-5, 6),
    main = paste0("Insulin - replicate ", i, " \n(", 
                  nrow(insulin_prots[[i]]), " proteins)"),
    mainCol = paste0(mycol, 70),
    outlineCol = darken(mycol))
  }

plot(NULL, xaxt = 'n', yaxt = 'n', bty = 'n', ylab = '',
     xlab='', xlim = 0:1, ylim = 0:1)
legend("topleft", legend = myleg, col = mycol, bty = "n",
       pch = 19, cex = 1, pt.cex = 1.6, ncol = 1)

# dev.off()
```


## t-SNE plots

Generate t-SNE plots of the data for each replicate

```{r, eval=TRUE}
## Basal map
tf_mirrorY <- c(TRUE, TRUE, FALSE)
tf_mirrorX <- c(FALSE, FALSE, TRUE)
tf_swt <- c(FALSE, FALSE, TRUE)

tsne_basal_prots <- vector("list", length(tf_mirrorX))
for (i in seq_along(basal_prots)) {
  set.seed(2)
  tsne_basal_prots[[i]] <- plot2D(basal_prots[[i]], method = "t-SNE",
                                  mirrorY = tf_mirrorY[i], mirrorX = tf_mirrorX[i], 
                                  axsSwitch = tf_swt[i], plot = FALSE)
  }

## Insulin map
tf_mirrorY <- c(FALSE, TRUE, FALSE)
tf_mirrorX <- c(TRUE, TRUE, TRUE)
tf_swt <- c(TRUE, FALSE, TRUE)
tsne_insulin_prots <- vector("list", length(tf_mirrorX))
for (i in seq_along(insulin_prots)) {
  set.seed(3)
  tsne_insulin_prots[[i]] <- plot2D(insulin_prots[[i]], method = "t-SNE",
                                   mirrorY = tf_mirrorY[i], mirrorX = tf_mirrorX[i],
                                   axsSwitch = tf_swt[i], plot = FALSE)
  }
```

Plot the data

```{r, eval=TRUE, fig.width=9, fig.height=9}
# png(file = here("2_data_visualisation/figures/Markers_tSNE_replicates.png"), height = 5000, width = 5000, units = "px", res = 500)

## Set colours and parameters for plots
orgclasses <- getMarkerClasses(basal_prots[[1]])
mycol <- getStockcol()[seq(orgclasses)]
myleg <- c(orgclasses, "Unknown")
mycolleg <- c(mycol, "grey")
  
par(mfrow = c(3, 3))
par(mar = c(5, 5, 4, 3))

## Basal
for (i in seq_along(basal_prots)) {
  prettyMap(
    tsne_basal_prots[[i]], 
    basal_prots[[i]], 
    fcol = "markers", 
    main = paste0("Basal - replicate ", i, " \n(", 
                  nrow(basal_prots[[i]]), " proteins)"),
    mainCol = paste0(mycol, 70),
    outlineCol = darken(mycol))
  }

## Insulin
for (i in seq_along(insulin_prots)) {
  prettyMap(
    tsne_insulin_prots[[i]], 
    insulin_prots[[i]], 
    fcol = "markers", 
    main = paste0("Insulin - replicate ", i, " \n(", 
                  nrow(insulin_prots[[i]]), " proteins)"),
    mainCol = paste0(mycol, 70),
    outlineCol = darken(mycol))
  }

plot(NULL, xaxt = 'n', yaxt = 'n', bty = 'n', ylab = '',
     xlab='', xlim = 0:1, ylim = 0:1)
legend("topleft", legend = myleg, col = mycol, bty = "n",
       pch = 19, cex = 1, pt.cex = 1.6, ncol = 1)

# dev.off()
```

# Individual Conditions (Basal, Insulin)

Load the data

```{r, eval=TRUE}
load(here("output/adipocyte_lopit_MSnSet_concat.rda"))
```

## PCA plots

Add markers and generate coordinates

```{r, eval=TRUE}
## Add markers
basal_p <- addMarkers(basal_p, markers = mrk)
insulin_p <- addMarkers(insulin_p, markers = mrk)

## Generate PCA coordinates
pca_basal_p <- plot2D(basal_p, method = "PCA", fcol = "markers", plot = FALSE)
pca_insulin_p <- plot2D(insulin_p, method = "PCA", fcol = "markers", plot = FALSE)
```

Now plot the data

```{r, eval=TRUE, fig.width=9, fig.height=3}
par(mfrow = c(1,3))
par(mar = c(5,6,4,1))
prettyMap(pca_basal_p, basal_p, fcol = "markers", 
          main = paste0("Basal ", "\n(", nrow(basal_p), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
prettyMap(pca_insulin_p, insulin_p, fcol = "markers", 
          main = paste0("Insulin ", "\n(", nrow(insulin_p), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
plot(NULL, xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
addLegend(basal_p, fcol = "markers", where = "left", cex = 1.4, ncol = 1, y.intersp = 1.1)
```


## t-SNE plots
```{r, eval=TRUE}
set.seed(1)
tsne_basal_p <- plot2D(basal_p, method = "t-SNE", fcol = "markers", plot = FALSE)
set.seed(1)
tsne_insulin_p <- plot2D(insulin_p, method = "t-SNE", fcol = "markers", plot = FALSE,
                         mirrorY = TRUE)
```

```{r, eval=TRUE, fig.width=9, fig.height=3}
# png(file = here("1_data_analysis/2_data_visualisation/figures/Markers_tSNE_concat.png"), width = 10000, height = 3500, units = "px", res =840)
par(mfrow = c(1,3))
par(mar = c(5,6,4,1))
prettyMap(tsne_basal_p, basal_p, fcol = "markers", 
          main = paste0("Basal ", "\n(", nrow(basal_p), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
prettyMap(tsne_insulin_p, insulin_p, fcol = "markers", 
          main = paste0("Insulin ", "\n(", nrow(insulin_p), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
plot(NULL, xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
addLegend(basal_p, fcol = "markers", where = "left", cex = 1.4, ncol = 1, y.intersp = 1.1)
# dev.off()
```


# Common proteins across all replictaes and all conditions

Load the data

```{r, eval=TRUE}
load(here("1_data_analysis/output/adipocyte_lopit_MSnSet_cmn.rda"))
```

## PCA plots

Add markers and generate PCA plots
```{r, eval=TRUE}
## Add markers
basal_cmn <- addMarkers(basal_cmn, markers = mrk)
insulin_cmn <- addMarkers(insulin_cmn, markers = mrk)

## Now generate PCAs
pca_basal_cmn <- plot2D(basal_cmn, method = "PCA", fcol = "markers", plot = FALSE)
pca_insulin_cmn <- plot2D(insulin_cmn, method = "PCA", fcol = "markers", plot = FALSE)
```

Plot the data
```{r, eval=TRUE, fig.width=9, fig.height=3}
par(mfrow = c(1,3))
par(mar = c(5,6,4,1))
prettyMap(pca_basal_cmn, basal_cmn, fcol = "markers", 
          main = paste0("Basal ", "\n(", nrow(basal_cmn), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
prettyMap(pca_insulin_cmn, insulin_cmn, fcol = "markers", 
          main = paste0("Insulin ", "\n(", nrow(insulin_cmn), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
plot(NULL, xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
addLegend(basal_cmn, fcol = "markers", where = "left", cex = 1.4, ncol = 1, y.intersp = 1.1)
```


## t-SNE plots

Generate t-SNE projections

```{r, eval=TRUE}
set.seed(1)
tsne_basal_cmn <- plot2D(basal_cmn, method = "t-SNE", fcol = "markers", plot = FALSE)
set.seed(1)
tsne_insulin_cmn <- plot2D(insulin_cmn, method = "t-SNE", fcol = "markers", plot = FALSE)
```

Plot the data

```{r, eval=TRUE, fig.width=9, fig.height=3}
par(mfrow = c(1,3))
par(mar = c(6,6,4,1))
prettyMap(tsne_basal_cmn, basal_cmn, fcol = "markers", 
          main = paste0("Basal ", "\n(", nrow(basal_cmn), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
prettyMap(tsne_insulin_cmn, insulin_cmn, fcol = "markers", 
          main = paste0("Insulin ", "\n(", nrow(insulin_cmn), " proteins)"),
          mainCol = paste0(mycol, 70), outlineCol = darken(mycol))
plot(NULL, xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
addLegend(basal_cmn, fcol = "markers", where = "left", cex = 1.4, ncol = 1, y.intersp = 1.1)
```
