---
title: "Notebook 4: Machine learning analysis"
author: "Olivia J. Conway"
date: "`r Sys.Date()`"
fontsize: 11pt
output:
  rmdformats::robobook:
    theme: default
    fig_caption: yes
    use_bookdown: yes
pdf_document:
    toc: yes
    number_sections: true
editor_options:
  chunk_output_type: console
---

# Overview
In this notebook we analyse the output machine learning results from running
the `bandle` classifier for differential localisation. We follow the [vignettes](https://www.bioconductor.org/packages/release/bioc/vignettes/bandle/inst/doc/v01-getting-started.html)
in the `bandle` package in Bioconductor.

## Install and load packages

Install R packages needed for data processing -

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
install.packages("BiocManager")
}

BiocManager::install(c("pRoloc",
                       "MSnbase",
                       "pRolocGUI",
                       "tidyverse",
                       "patchwork",
                       "gridExtra",
                       "colorspace",
                       "gplots",
                       "kableExtra",
                       "here",
                       "VennDiagram",
                       "eulerr",
                       "Rtsne",
                       "here"))
```

Load packages

```{r, eval=TRUE,message=FALSE, warning=FALSE}
library("pRoloc")
library("MSnbase")
library("pRolocGUI")
library("tidyverse")
library("patchwork")
library("gridExtra")
library("colorspace")
library("gplots")
library("kableExtra")
library("Rtsne")
library("here")
library("VennDiagram")
library("eulerr")
library("RColorBrewer")
library("bandle")
library("here")
```

## Load functions
```{r, eval=TRUE}
source(here("R/prettymap.R"))
source(here("R/bandle_outliers.R"))
```

# Processing BANDLE output

## Load BANDLE output

The `bandle` output can be found on Zenodo at DOI xxx as it is too big for
this repo.

```{r, eval=FALSE}
load(here("output/adipocyte_lopit_bandle_input.rda"))
load(here("output/adipocyte_lopit_bandle_output.rda"))
```

```{r, eval=TRUE, include=FALSE}
## Load the processed bandle output so we don't have to run in the notebook
load(here("output/adipocyte_lopit_bandle_input.rda"))
load(here("output/adipocyte_lopit_bandle_output.rda"))
load(here("output/adipocyte_lopit_bandle_processed.rda"))
```

```{r, eval=FALSE}
bandle_output <- bandleProcess(bandle_output)
```

## Assessing convergence

The `bandlea package used Markov chain Monte Carlo (MCMC) sampling. In MCMC
methods "chains" are drawn of random samples from the unknown posterior where
the current value depends on the previously drawn value (but not on values
before that). Once the chain has converged, its elements can be seen as a sample
from the target posterior probability distribution. A common first assessment of
whether chains have converged is to assess trace and density plots. We use the
`plotOutliers` function (which uses code from the [`coda` package](https://cran.r-project.org/web/packages/coda/index.html)) to generate
trace and density plots for each replicate and each condition to give us an idea
of convergence.

```{r, fig.height=10, fig.width=10}
plotOutliers(bandle_output)
```

We further examine the Gelman statistics to give us more information on convergence.

```{r}
calculateGelman(bandle_output)
```

The trace and density plots look as expected and the Gelman statistics are all
< 1.2, thus we take all 4 chains forward for our analysis.

<!-- ## Extract the posterior estimates and protein allocation predictions -->

<!-- Following the vignettes in the `bandle` package we extract the posterior -->
<!-- estimates and then visualise the distribution of protein allocations prior -->
<!-- to any filtering. -->

<!-- ```{r, eval=FALSE} -->
<!-- pe1 <- summaries(bandle_output)[[1]]@posteriorEstimates -->
<!-- pe2 <- summaries(bandle_output)[[2]]@posteriorEstimates -->

<!-- barplot(table(pe1$bandle.allocation), col = getStockcol()[2], -->
<!--         las = 2, ylim = c(0,1200), main = "Basal: Protein allocation \n pre-thresholding", -->
<!--         ylab = "Number of proteins") -->
<!-- barplot(table(pe2$bandle.allocation), col = getStockcol()[2], -->
<!--         las = 2, ylim = c(0,1200), main = "Insulin: Protein allocation \n pre-thresholding", -->
<!--         ylab = "Number of proteins") -->
<!-- ``` -->

<!-- **Glossary:** -->

<!-- *bandle.allocation:* contains the the localisation predictions to one of the -->
<!-- subcellular classes that appear in the training data.   -->
<!-- <br> -->
<!-- *bandle.probability:* the allocation probability, corresponding to the mean -->
<!-- of the distribution probability.   -->
<!-- <br> -->
<!-- *bandle.outlier:* the probability of being an outlier. A high value -->
<!-- indicates that the protein is unlikely to belong to any annotated class (and is -->
<!-- hence considered an outlier).   -->
<!-- <br> -->
<!-- *bandle.probability.lowerquantile* and *bandle.probability.upperquantile:* -->
<!-- the upper and lower quantiles of the allocation probability distribution.   -->
<!-- <br> -->
<!-- *bandle.mean.shannon:* the Shannon entropy, measuring the uncertainty in the -->
<!-- allocations (a high value representing high uncertainty; the highest value is -->
<!-- the natural logarithm of the number of classes).   -->
<!-- <br> -->
<!-- *bandle.differential.localisation:* the differential localisation probability.   -->
<!-- <br> -->

## Extracting the bandle results

We use `bandlePredict` to append results to original `MSnSet.` Note, results are 
appended to **first** replicate of each condition in the `MSnSet` data structure.

```{r, eval=TRUE}
res <- bandlePredict(objectCond1 =  basal, 
                     objectCond2 =  insulin, 
                     params = bandle_output, 
                     fcol = "markers")

res_basal <- res[[1]]
res_insulin <- res[[2]]
```


```{r, eval=TRUE}
# Update labels of each replicate
for (i in seq(res_basal)) {
  res_basal[[i]] <- updateFvarLabels(res_basal[[i]],
                                     label = paste0("rep", i),
                                     sep = "_")
  res_insulin[[i]] <- updateFvarLabels(res_insulin[[i]],
                                       label = paste0("rep", i),
                                       sep = "_")
}

# Combine into MSnSet
res_basal_p <- MSnbase::combine(res_basal[[1]], res_basal[[2]])
res_basal_p <- MSnbase::combine(res_basal_p, res_basal[[3]])
res_insulin_p <- MSnbase::combine(res_insulin[[1]], res_insulin[[2]])
res_insulin_p <- MSnbase::combine(res_insulin_p, res_insulin[[3]])
```


# Exploring the allocations and outliers

BANDLE gives us the `allocation probability` (the mean posterior probability for
the master protein subcellular allocations computed by TAGM-MCMC), and the
`outlier probability` (the posterior probability for the protein to belong to
the outlier component rather than any of the annotated components in the basal
LOPIT data).

Below we threshold on both of these values, to remove allocations from proteins
that have a low allocation probability or high probability of being an outlier
to make our allocations more robust and reduce over-allocation to subcellular
compartments. Proteins where the allocation or outlier probability value does
not meet the threshold are left unannotated and given "unknown" localisation.

## Basal results

```{r, eval=TRUE}
## get proteins NOT used as markers
basal_un_msnset <- unknownMSnSet(res_basal_p, fcol = "markers_rep1")

## plot BANDLE allocation probability
nt_prob_b <- tapply(fData(basal_un_msnset)[, "bandle.probability_rep1"],
                    fData(basal_un_msnset)[, "bandle.allocation_rep1"], 
                    summary)
boxplot(nt_prob_b, las = 2, main = "BANDLE probabilities by organelle")
```

The closer the outlier probability is to 1, the more likely it is we have
outliers for that organelle class

```{r, eval=TRUE}
nt_out_b <- tapply(fData(basal_un_msnset)[, "bandle.outlier_rep1"], 
                 fData(basal_un_msnset)[, "bandle.allocation_rep1"], 
                 summary)
boxplot(nt_out_b, las = 2, main = "BANDLE outlier component prob. by organelle")
```

Here we create a new column where we invert outlier probability (so the closer
the outlier probability is to 1, the more likely it is we have outliers for that
organelle class).

```{r, eval=TRUE}
## create a new column for the outlier probability
fData(res_basal_p)$outlier.prob <- 1 - fData(res_basal_p)$bandle.outlier_rep1
fData(basal_un_msnset)$outlier.prob <- 1 - fData(basal_un_msnset)$bandle.outlier_rep1

nt_out_b <- tapply(fData(basal_un_msnset)[, "outlier.prob"], 
                 fData(basal_un_msnset)[, "bandle.allocation_rep1"], 
                 summary)
# boxplot(nt_out_b, las = 2, main = "BANDLE (1 - outlier) component prob. by organelle: Basal")
```


### Thresholding on allocation probability only ("low confidence")

Fist we threshold on allocation probability.
<br>
<br>
`fcol:` name of the prediction column in the feature data  
`scol:` name of the prediction score column  
`mcol:` name of the feature meta data containing the labelled training data (doesn't have bandle allocation/probability scores)  
`t:` the score threshold, everything with score <t are set to 'unknown'

```{r, eval=TRUE}
# here we set the allocation to "unknown" for all proteins where the allocation probability is <0.9)
res_basal_p <- getPredictions(res_basal_p, 
                              fcol = "bandle.allocation_rep1", 
                              scol = "bandle.probability_rep1",
                              mcol = "markers_rep1",
                              t = 0.9)
```


### Thresholding on allocation and outlier probability ("high confidence")

As the outlier probability varies a lot across organelles, we threshold each one
individually, based on it's median outlier probability score.


```{r, eval=TRUE}
summary_df_b <- data.frame(
  Organelle = names(nt_out_b),
  Min = sapply(nt_out_b, function(x) x["Min."]),
  Q1 = sapply(nt_out_b, function(x) x["1st Qu."]),
  Median = sapply(nt_out_b, function(x) x["Median"]),
  Mean = sapply(nt_out_b, function(x) x["Mean"]),
  Q3 = sapply(nt_out_b, function(x) x["3rd Qu."]),
  Max = sapply(nt_out_b, function(x) x["Max."]),
  row.names = NULL
)
print(summary_df_b)
```

```{r}
t_user <- c(0, 0, 0, 0, 0, 0.5, 0, 0, 0, 0.5, 0)
names(t_user) <- c(getMarkerClasses(res_basal_p, fcol = "markers_rep1"), "unknown")

res_basal_p <- getPredictions(res_basal_p, 
                              fcol = "bandle.allocation_rep1.pred", 
                              scol = "outlier.prob",
                              mcol = "markers_rep1",
                              t = t_user)
```

```{r}
# if median outlier prob = 0 then threshold = 1, else threshold = median
## proteins allocated to an organelle with an outlier probability of <threshold are changed to "unknown"
t_user <- c(0.7479, 1, 0.1470, 1, 1, 1, 1, 0.0102, 1, 1, 0)
names(t_user) <- c(getMarkerClasses(res_basal_p, fcol = "markers_rep1"), "unknown")

res_basal_p <- getPredictions(res_basal_p, 
                     fcol = "bandle.allocation_rep1.pred.pred", 
                     scol = "outlier.prob",
                     mcol = "markers_rep1",
                     t = t_user)
```

### Basal: Visualisation of the allocation results

```{r, eval=TRUE, fig.width=8, fig.height=10}
pca_res_basal_p <- plot2D(res_basal_p, fcol = "markers_rep1", plot = FALSE)

par(mfrow = c(3,2))
par(mar = c(6, 6, 6, 2))
prettyMap_overlay(pca_res_basal_p, res_basal_p, 
                  main = "Basal: markers", 
                  fcol = "markers_rep1")
prettyMap_overlay(pca_res_basal_p, res_basal_p, 
                  main = "Basal: bandle protein allocation", 
                  fcol = "bandle.allocation_rep1")
prettyMap_overlay(pca_res_basal_p, res_basal_p, 
                  main = "Basal: 'low' outlier threshold", 
                  fcol = "bandle.allocation_rep1.pred")
prettyMap_overlay(pca_res_basal_p, res_basal_p, 
                  main = "Basal: 'medium' outlier threshold", 
                  fcol = "bandle.allocation_rep1.pred.pred")
prettyMap_overlay(pca_res_basal_p, res_basal_p, 
                  main = "Basal: 'high' outlier threshold", 
                  fcol = "bandle.allocation_rep1.pred.pred.pred")

plot(NULL, xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
addLegend(res_basal_p, fcol = "markers_rep1", where = "center", cex = 1, ncol = 2, y.intersp = 1.2)
```


## Insulin results

```{r, eval=TRUE}
## get proteins NOT used as markers
insulin_un_msnset <- unknownMSnSet(res_insulin_p, fcol = "markers_rep1")

## plot BANDLE allocation probability
nt_prob_ins <- tapply(fData(insulin_un_msnset)[, "bandle.probability_rep1"], 
             fData(insulin_un_msnset)[, "bandle.allocation_rep1"], 
             summary)
boxplot(nt_prob_ins, las = 2, main = "BANDLE probabilities by organelle")
```

The closer the outlier probability is to 1, the more likely it is we have
outliers for that organelle class

```{r, eval=TRUE}
nt_out_ins <- tapply(fData(insulin_un_msnset)[, "bandle.outlier_rep1"], 
                 fData(insulin_un_msnset)[, "bandle.allocation_rep1"], 
                 summary)
boxplot(nt_out_ins, las = 2, main = "BANDLE outlier component prob. by organelle")
```


Again, we create a new column where we invert outlier probability (so the closer
the outlier probability is to 1, the more likely it is we have outliers for that
organelle class). This allows us to use the `getPredictions` function to
threshold on this value.

```{r, eval=TRUE}
## create a new column for the outlier probability
fData(res_insulin_p)$outlier.prob <- 1 - fData(res_insulin_p)$bandle.outlier_rep1
fData(insulin_un_msnset)$outlier.prob <- 1 - fData(insulin_un_msnset)$bandle.outlier_rep1

## re-plot outlier probability (now the closer to 0 the more likely we have outliers for this class)
nt_out_ins <- tapply(fData(insulin_un_msnset)[, "outlier.prob"], 
                 fData(insulin_un_msnset)[, "bandle.allocation_rep1"], 
                 summary)
```

### Thresholding on allocation probability only ("low confidence")

Fist we threshold on allocation probability.
<br>
<br>
`fcol:` name of the prediction column in the feature data  
`scol:` name of the prediction score column  
`mcol:` name of the feature meta data containing the labelled training data (doesn't have bandle allocation/probability scores)  
`t:` the score threshold, everything with score <t are set to 'unknown'
```{r, eval=TRUE}
# here we set the allocation to "unknown" for all proteins where the allocation probability is <0.9)
res_insulin_p <- getPredictions(res_insulin_p, 
                     fcol = "bandle.allocation_rep1", 
                     scol = "bandle.probability_rep1",
                     mcol = "markers_rep1",
                     t = 0.9)
```

### Thresholding on allocation and outlier probability ("high confidence")

As the outlier probability varies a lot across organelles, we threshold each one
individually, based on it's median outlier probability score.

```{r, eval=FALSE}
summary_df_ins <- data.frame(
  Organelle = names(nt_out_ins),
  Min = sapply(nt_out_ins, function(x) x["Min."]),
  Q1 = sapply(nt_out_ins, function(x) x["1st Qu."]),
  Median = sapply(nt_out_ins, function(x) x["Median"]),
  Mean = sapply(nt_out_ins, function(x) x["Mean"]),
  Q3 = sapply(nt_out_ins, function(x) x["3rd Qu."]),
  Max = sapply(nt_out_ins, function(x) x["Max."]),
  row.names = NULL
)
print(summary_df_ins)
```

```{r, eval=TRUE}
# if median outlier prob = 0 (highest chance of having outliers) then threshold = 0.5, else threshold = 0
## so in organelles with a high chance of having outliers, we remove allocations from proteins in these organelles which have a >50% chance of being an outlier.
### proteins allocated to Nucleus with an outlier probability of <0.5 are changed to "unknown"
t_user_ins <- c(0, 0, 0, 0, 0, 0.5, 0, 0, 0, 0, 0)
names(t_user_ins) <- c(getMarkerClasses(res_insulin_p, fcol = "markers_rep1"), "unknown")
res_insulin_p <- getPredictions(res_insulin_p, 
                     fcol = "bandle.allocation_rep1.pred", 
                     scol = "outlier.prob",
                     mcol = "markers_rep1",
                     t = t_user_ins)
```

```{r, eval=TRUE}
# if median outlier prob = 0 then threshold = 1, else threshold = median
## proteins allocated to an organelle with an outlier probability of <threshold are changed to "unknown"
t_user_ins <- c(1, 1, 1, 1, 1, 1, 1, 0.9014, 1, 0.0372, 0)
names(t_user_ins) <- c(getMarkerClasses(res_insulin_p, fcol = "markers_rep1"), "unknown")
res_insulin_p <- getPredictions(res_insulin_p, 
                     fcol = "bandle.allocation_rep1.pred.pred", 
                     scol = "outlier.prob",
                     mcol = "markers_rep1",
                     t = t_user_ins)
```

### Insulin: Visualisation of the allocation results

```{r, eval=TRUE, fig.width=8, fig.height=10}
pca_res_insulin_p <- plot2D(res_insulin_p, fcol = "markers_rep1", plot = FALSE)

par(mfrow = c(3, 2))
par(mar = c(6, 6, 6, 2))
prettyMap_overlay(pca_res_insulin_p, res_insulin_p, 
                  main = "Insulin: markers", 
                  fcol = "markers_rep1")
prettyMap_overlay(pca_res_insulin_p, res_insulin_p, 
                  main = "Insulin: bandle protein allocation", 
                  fcol = "bandle.allocation_rep1")
prettyMap_overlay(pca_res_insulin_p, res_insulin_p, 
                  main = "Insulin: 'low' outlier threshold", 
                  fcol = "bandle.allocation_rep1.pred")
prettyMap_overlay(pca_res_insulin_p, res_insulin_p, 
                  main = "Insulin: 'medium' outlier threshold", 
                  fcol = "bandle.allocation_rep1.pred.pred")
prettyMap_overlay(pca_res_insulin_p, res_insulin_p, 
                  main = "Insulin: 'high' outlier threshold", 
                  fcol = "bandle.allocation_rep1.pred.pred.pred")
```

<!-- # Save final MSnSet object for use in later notebooks -->
<!-- ```{r, eval=FALSE} -->
<!-- save(res_basal_p, res_insulin_p, -->
<!--      file = here("output/adipocyte_lopit_allocations.rda")) -->
<!-- ``` -->

# Differential localisation

The differential localisation probability tells us which proteins are most
likely to *differentially localise*, and therefore exhibit a change in their
steady-state subcellular location. The differential localisation probability is
found in the `bandle.differential.localisation` column of the
`bandleParams`output.

```{r numtransloc, eval=TRUE}
# extract all the diff loc probabilities
pe1 <- summaries(bandle_output)[[1]]@posteriorEstimates
pe2 <- summaries(bandle_output)[[2]]@posteriorEstimates
diffloc_probs <- pe1$bandle.differential.localisation
```

```{r, eval=TRUE, fig.width=8, fig.height=6}
par(mar = c(5, 5, 6, 2))
plot(diffloc_probs[order(diffloc_probs, decreasing = TRUE)],
     col = getStockcol()[2], pch = 1, 
     ylab = "Differential localisation probability",
     xlab = "Protein rank", 
     main = "Differential localisation rank plot", 
     xlim = c(0,3500), ylim = c(0,1))
```

We can examine the number of differentially localising proteins we have 
according to probability. 

```{r, eval=TRUE}
# Define probability threshold
thresholds <- c(1, .95, seq(.9, .1, by = -.1))

# Count proteins meeting each threshold
counts <- sapply(thresholds, function(threshold) sum(diffloc_probs >= threshold))

# Print the result
result_df <- data.frame(Threshold = thresholds, Count = counts)
print(result_df)
```

## Candidates

We define all proteins with a differential localisation probability = 1 as candidates.

```{r}
ind <- which(fData(res_basal_p)$bandle.differential.localisation_rep1 == 1)
candidates <- featureNames(res_basal_p)[ind]
print(candidates)
```

## Visualisation

```{r}
bandle_res_basal <- MSnSetList(list(res_basal_p[, 1:10],
                                    res_basal_p[, 2:20],
                                    res_basal_p[, 21:30])) 

bandle_res_insulin <- MSnSetList(list(res_insulin_p[, 1:10],
                                      res_insulin_p[, 2:20],
                                      res_insulin_p[, 21:30])) 

orgs <- c(union(getMarkerClasses(bandle_res_basal[[1]], "bandle.allocation_rep1.pred.pred.pred"), 
                getMarkerClasses(bandle_res_insulin[[1]], "bandle.allocation_rep1.pred.pred.pred")))

circos_cols <- c(getStockcol()[seq_along(orgs)], "grey")
(colscheme <- setNames(circos_cols, c(orgs, "unknown"))) 

subset_msnset <- list(bandle_res_basal[[1]][ind, ], 
                      bandle_res_insulin[[1]][ind, ])
```


Chord diagram

```{r, fig.width=10, fig.height=10}
plotTranslocations(subset_msnset, fcol = "bandle.allocation_rep1.pred.pred.pred", col = colscheme, type = "chord")
```

Alluvial

```{r, fig.height=11, fig.width=8}
plotTranslocations(subset_msnset, fcol = "bandle.allocation_rep1.pred.pred.pred", col = colscheme, type = "alluvial")
```


Table

```{r}
plotTable(subset_msnset, fcol = "bandle.allocation_rep1.pred.pred.pred", all = TRUE)
```

 